Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-12-01 Documents/Paper/ms.tex
Going through the introduction to add more background info

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
components in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. A common statistical example is finite mixture
modelling. In the Bayesian framework, these problems are typically solved using
one of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.
Both of the examples demonstrate phase transitions in the relationship between
the likelihood and the cumulative prior mass.}

\section{Introduction}\label{sec:introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Consider the following class of inference problems. There is some unknown
number, $N$, of components in a (physical or metaphorical) region. Each of the $N$ components has a property
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a two-dimensional position and a mass). These problems are often
challenging to solve due to the unknown (and potentially large) dimensionality.
In addition, they also have the so-called ``label-switching degeneracy'' issue
\citep{label_switching}, where the meaning of the model is invariant to
switching the identities of the components. Finite mixture modelling is a common
statistical example.

To carry out the inference, we must assign a prior to the components' properties
$\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.
The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

@@

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the interim prior for $\{\xx_i\}$ is iid and does not depend on $N$.
Given data $\mathcal{D}$, we usually want to calculate properties of the
posterior distribution, given by:
\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\} | \mathcal{D}) \propto
p(N, \hyperparams, \{\xx_i\})
p(\mathcal{D} | N, \hyperparams, \{\xx_i\})\label{eq:bayes}
\end{eqnarray}
The structure of the model is depicted graphically in Figure~\ref{fig:pgm}.

@@

In this paper, we introduce an approach that uses trans-dimensional MCMC moves
to infer $N$, but embeds this process in the Nested Sampling framework to
overcome difficulties that would be encountered by sampling the posterior
distribution directly. Our motivation
for using Nested Sampling is not primarily to calculate the marginal
likelihood $\mathcal{Z}$ (including summing over $N$, as in
Equation~\ref{eq:evidence}),
although $\mathcal{Z}$ is readily available from the output.
The motivation for using Nested Sampling, rather than just sampling the
posterior using trans-dimensional MCMC, is that the posterior often contains
difficult features, such as strong dependencies or multiple modes, that cause
problems with mixing. Diffusive Nested Sampling
\citep[DNS][]{dnest} replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained versions of the prior, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of annealed importance sampling
\citep{neal} or parallel tempering \citep{pt},
where the target distribution is modified from the posterior to something
``easier''. However,
the sequence of distributions used in Nested Sampling avoids some of the
problems with annealing-based methods. In particular, there is no need to choose
a temperature schedule (a potentially large set of tuning parameters),
and the method does not fail when phase transitions
are present.

>>>>

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
components in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. A common statistical example is finite mixture
modelling. In the Bayesian framework, these problems are typically solved using
one of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.
Both of the examples demonstrate phase transitions in the relationship between
the likelihood and the cumulative prior mass, highlighting the need for
Nested Sampling.}

\section{Introduction}\label{sec:introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Consider the following class of inference problems. There is some unknown
number, $N$, of components in a (physical or metaphorical) region. Each of the $N$ components has parameters
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a two-dimensional position and a mass). These problems are often
challenging to solve due to the unknown (and potentially large) dimensionality.
In addition, they also have the so-called ``label-switching degeneracy'' issue
\citep{label_switching}, where the meaning of the model is invariant to
switching the identities of the components. Finite mixture modelling is a common
statistical example.

To carry out the inference, we must assign a prior to the components' parameters
$\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``conditional prior''
given some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.
The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

@@

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the conditional prior for $\{\xx_i\}$ is iid and does not depend on $N$.
Given data $\mathcal{D}$, we usually want to calculate properties of the
posterior distribution, given by:
\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\} | \mathcal{D}) \propto
p(N, \hyperparams, \{\xx_i\})
p(\mathcal{D} | N, \hyperparams, \{\xx_i\})\label{eq:bayes}
\end{eqnarray}
The structure of the model is depicted graphically in Figure~\ref{fig:pgm}.

@@

In this paper, we introduce an approach that uses trans-dimensional MCMC moves
to infer $N$, but embeds this process in the Nested Sampling framework to
overcome difficulties that would be encountered by sampling the posterior
distribution directly. Our motivation
for using Nested Sampling is not primarily to calculate the marginal
likelihood $\mathcal{Z}$ (including summing over $N$, as in
Equation~\ref{eq:evidence}),
although $\mathcal{Z}$ is readily available from the output.
The motivation for using Nested Sampling, rather than just sampling the
posterior using trans-dimensional MCMC, is that the posterior often contains
difficult features, such as strong dependencies or multiple modes, that cause
problems with mixing.

In addition to these well-known challenges,
{\it phase transitions} \citep{skilling} can also cause difficulties for
Bayesian computation. Phase transitions occur when the posterior distribution
is a mixture of a ``slab'' with high volume but low density, and a ``spike''
with low volume but high density. To understand why this causes problems,
imagine sampling the posterior using the Metropolis algorithm. If the sampler
is in the slab, it will only rarely propose a move into the spike, because the
spike's volume is so small. If the sampler is in the spike, a proposal to move
into the slab will have a very low acceptance probability because the density
of the slab is so low relative to that of the spike.

Diffusive Nested Sampling
\citep[DNS][]{dnest} replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained versions of the prior, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of annealed importance sampling
\citep{neal} or parallel tempering \citep{pt},
where the target distribution is modified from the posterior to something
``easier''. However,
the sequence of distributions used in Nested Sampling avoids some of the
problems with annealing-based methods. In particular, there is no need to choose
a temperature schedule (a potentially large set of tuning parameters),
and the method does not fail when phase transitions
are present.
